---
title: "Speed and Accuracy Tests"
subtitle: "Package 'SunCalcMeeus' `r packageVersion('SunCalcMeeus')` "
author: "Pedro J. Aphalo"
date: "`r packageDate('SunCalcMeeus')`"
output: 
  rmarkdown::html_document:
    toc: yes
---

## Introduction

R package 'SunCalcMeeus' collects functions related to the computation of
the position of the sun and times, including day length and night length, from
package 'photobiology'. It also includes function for computing empirical
estimates of the relative air mass (AM) for different sun elevation angles.

The position of the sun is computed using Meeus well known algorithm, which is
accurate over a broad range of dates both into the past and future. The
algorithm is described as useful for dates between 1000 BC to 3000 AD, however
accuracy of the implementation in 'SunCalcMeeus' could be affected by how Julian
days are computed and the limited precision of the computer representation of
real numbers as double precision floating point numbers. Results for years 1700
to 2100 are checked below against the tables of the US Naval Observatory as
reference.

**Note:** Although for many uses, including biology, energy use and history research and
applications, Meeus algorithm is considered very accurate, for use in astronomy,
it is considered not the most accurate. One collection of libraries for
astronomy, in FORTRAN and ANSI C versions, is provided by the International
Astronomy Union under the name of Standards of Fundametal Astronomy
[SOFA](http://www.iausofa.org/). The [NASA web page JPL
Horizons](https://ssd.jpl.nasa.gov/) provides on-line computed up-to-date
ephemerides for for huge number of celestial bodies covering from 9999 BC to
9999 AD.

A comparison to package 'suncalc' is also done both for performance and accuracy.
This is of interest as the two packages are likely to differ both in performance
and accuracy, and their use could preferable in different situations.


```{r, message=FALSE}
library(SunCalcMeeus)
library(lubridate)
library(dplyr)
library(tidyr)
library(readr)
library(microbenchmark)
library(suncalc)
library(ggplot2)
library(ggpp)
library(ggdensity)
```

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4)
```

```{r, eval=TRUE, include=FALSE}
options(tibble.print_max = 4, tibble.print_min = 3)
```

## Accuracy

### Sunset and sunrise times

In this section I check the values from 'SunCalcMeeus' against the on-line data
from the [Astronomical Applications Department of the U.S. Naval
Observatory](https://aa.usno.navy.mil/data/index). For any comparison of sunrise
or sunset times, it is important to ensure that the twilight definition used is
the same. In other words, at what instant in time the start and end of the day
are assumed to be is crucial. In addition, if an atmospheric refraction
correction is applied or not must also be consistent. The astronomical
definition of sunrise and sunset is when the upper rim of the solar disk is at
the horizon, and most frequently a refraction correction is applied when the
observer is on Earth. This corresponds to `twilight = "sunlight"` in the
functions from 'SunCalcMeeus'.

The format of the table downloaded, or more precisely copied and paste from the
web page, is such that converting the data into a shape compatible with that
used by 'SunCalcMeeus' takes a some effort. I use for the comparison the
geographic location of Greenwhich, U.K., and the GMT time zone.

```{r, warning=FALSE}
# map
vars2months <- rep(month.abb, each = 2)
names(vars2months) <- paste("X", 2L:25L, sep = "")

vars2rise_or_set <- rep(c("sunrise", "sunset"), times = 12)
names(vars2rise_or_set) <- paste("X", 2L:25L, sep = "")

for (this.year in c(1700, 1900, 2000, 2100)) {
  read_fwf(paste("data/USNO-", this.year, ".txt", sep = ""), skip = 9, n_max = 31, 
           col_types = "ccccccccccccccccccccccccc") |>
    pivot_longer(X2:X25, values_to = "time", cols_vary = "slowest") |>
    mutate(year = this.year,
           month = vars2months[name],
           day = X1,
           event = vars2rise_or_set[name],
           X1 = NULL,
           name = NULL,
           hhmm = ymd_hm(paste(year, "-", month, "-", day, " ", time, sep = ""),
                         tz = "UTC"),
           day = ymd(paste(year, "-", month, "-", day, sep = ""),
                     tz = "UTC"),
           time = NULL)|>
    select(day, event, hhmm) -> all.data
  
  sunrises.df <- subset(all.data, event == "sunrise", select = -2)
  names(sunrises.df)[2] <- "usno.sunrise"
  sunsets.df <- subset(all.data, event == "sunset")
  sunrises.df$usno.sunset <- sunsets.df$hhmm
  usno.df <- na.omit(sunrises.df)
  
  meeus_times.df <- day_night(usno.df$day, unit.out = "datetime", twilight = "sunlight")
  
  full_join(meeus_times.df, usno.df, by = join_by(day)) |>
    mutate(sunrise.diff = sunrise - usno.sunrise,
           sunset.diff = sunset - usno.sunset) -> temp.df
  
  assign(paste("times_", this.year, ".df", sep = ""), temp.df)
}  

times_all.df <- bind_rows(times_1700.df, times_1900.df, times_2000.df, times_2100.df)
times_all.df$year <- year(times_all.df$day)
```

The quantiles for the pooled differences from all years tested are:

```{r}
quantile(times_all.df$sunrise.diff)
```

```{r}
quantile(times_all.df$sunset.diff)
```

Differences were in all cases less than $\pm 71$ s, while the resolution of the
data from the U.S. Naval Observatory is 1 min. This good match was true for
years 1700, 1900, 2000 and 2100, and for both sunrise and sunset times. So,
Meeus algorithm is generating values differing little from known-good values.

We plot the differences for these four years, both for sunrise and sunset.
First as a density distribution and farther down in a scatter plots.

```{r, message=FALSE}
ggplot(times_all.df, aes(sunrise.diff)) +
  stat_density() +
  facet_wrap(facets = vars(year), labeller = label_both) +
  expand_limits(x = c(-80, 80)) +
  labs(x = "Difference in sunrise time (s)")
```

```{r, message=FALSE}
ggplot(times_all.df, aes(sunset.diff)) +
  stat_density() +
  facet_wrap(facets = vars(year), labeller = label_both) +
  expand_limits(x = c(-80, 80)) +
  labs(x = "Difference in sunset time (s)")
```

```{r, message=FALSE}
ggplot(times_all.df, aes(day, sunrise.diff)) +
  geom_point() +
  labs(y = "Difference in sunrise time (s)", x = "Date") +
  scale_x_datetime(date_labels = "%b") +
  facet_wrap(facets = vars(year), scales = "free_x", labeller = label_both)
```

```{r, message=FALSE}
ggplot(times_all.df, aes(day, sunset.diff)) +
  geom_point() +
  labs(y = "Difference in sunset time (s)", x = "Date") +
  scale_x_datetime(date_labels = "%b") +
  facet_wrap(facets = vars(year), scales = "free_x", labeller = label_both)
```

'SunCalcMeeus' provides accurate estimates, but how does its performance
compare to 'suncalc'? Is the possibly slower performance worth the expected
improvement in accuracy?

## Performance

Meeus mathematical algorithm is more computationally intensive than other less
accurate ones, such as that implemented in package 'suncalc'. However, I have
optimized the implementation in 'SunCalcMeeus' for the case of long
vectors of instants in time or of dates at each geographic location.

**Note:** The equations used in 'suncalc' are "mostly" from the web page [Astronomy
Answers](https://www.aa.quae.nl/en/reken/zonpositie.html). I was unable to find
a published source for the algorithm other than the web page. It is not clear
from the documentation if an atmospheric refraction correction is applied or
not.

### Sun position

To check the call overhead, I start by benchmarking the computation of the sun
position at a single instant in time at a single geographic location.

```{r}
time <- now()
```

```{r, warning=FALSE}
microbenchmark(sun_angles(time), 
               getSunlightPosition(date = time, lat = 51.5, lon = 0), 
               times = 500L,
               unit = "millisecond")
```

I next test the performance of the computation of the position of the sun with 'SunCalcMeeus' using 10 000 instants in time stored in a `POSIXct` vector and nine geographic locations in a `data.frame` with columns for longitude and latitude and rows for geographic locations.

```{r}
times <- ymd_hms("2024-12-11 00:00:00") + minutes(1:1e4)
locations <- data.frame(lon = c(0, 0, 0, 90, 90, 90, -90, -90, -90),
                        lat = c(0, 30, 60, 0, 30, 60, 0, 30, 60))
range(times)
```

We benchmark performance by evaluating the statement 20 times. The median time
is the most informative.

```{r}
microbenchmark(sun_angles(times, geocode = locations), 
               times = 20L)
```

Next we will test the performance with 1 000 000 instants in time and one 
geographic location, the default one (Greewich, UK).

```{r}
times <- ymd_hms("2024-12-11 00:00:00") + minutes(1:1e6)
range(times)
```
There is one problem, possibly a division by zero, affecting azimuth for one
instant in time.

```{r}
sun_angles.df <- sun_angles(times)
sapply(sun_angles.df, anyNA)
which(is.na(sun_angles.df$azimuth))
sun_angles.df[which(is.na(sun_angles.df$azimuth)), ]
```

We run the benchmarks with the longer vector of instants in time:

```{r, warning=FALSE}
microbenchmark(sun_angles(times), 
               getSunlightPosition(date = times, lat = 51.5, lon = 0), 
               times = 5L,
               unit = "second")
```

`sun_angles()` takes between 2.5 and 3.2 times as much time to run than
`getSunlightPosition()` but returns six computed values while
`getSunlightPosition` returns only two. For most uses they are both most likely
fast enough.

### Sunrise and sunset times

Checking performance as above, but day by day for 600 years.

```{r, warning=FALSE}
dates <- ymd_hms("1700-01-01 00:00:00", tz = "UTC") + days(0:(600 * 365 + 145)) 
length(dates)
range(dates)
```

First just a single date to asses the call overhead.

```{r, warning=FALSE}
microbenchmark(day_night(dates[1]), 
               day_night(dates[1], unit.out = "datetime"), 
               getSunlightPosition(date = dates[1], lat = 51.5, lon = 0), 
               times = 100L,
               unit = "millisecond")
```

And with every day in 600 years.

```{r, warning=FALSE}
microbenchmark(day_night(dates), 
               day_night(dates, unit.out = "datetime"), 
               getSunlightPosition(date = dates, lat = 51.5, lon = 0), 
               times = 10L,
               unit = "second")
```

The difference is similar to that above for the position. But once again, for
most uses they are both fast enough.

```{r}
meeus_dn.df <- day_night(dates[1], tz = "UTC", unit.out = "datetime", twilight = "sunlight")
colnames(meeus_dn.df)
```

Now we use the 'suncalc' package to compute values for the same day.

```{r}
suncalc_dn.df <- getSunlightTimes(date = as.Date(dates[1], tz = "UTC"), lat = 51.5, lon = 0)
colnames(suncalc_dn.df)
```

## Consistency between 'suncalc' and 'SunCalcMeeus'

### Sun position

The next question is do the values returned by functions from both packages
agree? and does the agreement hold for past and future centuries. Let's check
every hour for 600 years starting from 1 January 1700.

```{r, warning=FALSE}
times <- ymd_hms("1700-01-01 00:00:00", tz = "UTC") + hours(1:5.25948e6)
range(times)
```

```{r, warning=FALSE}
meeus.df <- sun_angles(times, use.refraction = TRUE)
meeus.df
```

Now we use the 'suncalc' package to compute the same values.

```{r}
suncalc.df <- getSunlightPosition(date = times, lat = 51.5, lon = 0)
suncalc.df$elevation <- suncalc.df$altitude * 180 / pi # radians to degrees
suncalc.df$azimuth <- suncalc.df$azimuth * 180 / pi # radians to degrees
suncalc.df$azimuth <- suncalc.df$azimuth + 180 # radians to degrees
head(suncalc.df, n = 3L)
```

We compute the differences in sun elevations, and their range. Over this
extended period, the maximum and minimum differences are substantial. We plot a
smooth spline on the absolute value of the differences, as from the range we
already know that they maxima and minimum differences are of opposite signs. The
sun elevation values from the two functions are closest during the first half of
the 21st century. On the other hand we see that the fitted spline does not reach
the extreme values by a large margin. What is going on?

```{r, message=FALSE}
differences <- suncalc.df$elevation - meeus.df$elevation
differences.range <- range(differences)
differences.range

ggplot(data.frame(times, differences), aes(times, abs(differences))) +
  stat_smooth(method = "gam") +
  stat_panel_counts(label.x = "centre") +
  labs(x = "Time", y = "Absolute difference in sun elevation (degrees)") +
  expand_limits(y = 0)
```

If we plot the 2D density distribution we get a very surprising pattern. So as
we move into the past we mainly have thre concentrations of differences:
positive, no difference or negatine, with fewer intermediate values. When we
move into the future we get four rays of concentrated difference values.

```{r}
ggplot(data.frame(times, differences), aes(times, differences)) +
  stat_hdr() +
  stat_panel_counts(label.x = "centre") +
  labs(x = "Time", y = "Difference in sun elevation (degrees)")
```

A difference in sun elevation of 3.5 degrees is too much for many applications.
So which of the two algorithms is more accurate? Before checking this, let's
check time related quantities.

### Sunrise and sunset times

Lets check every day for 600 years starting from 1 January 1700.

```{r, warning=FALSE}
times <- ymd_hms("1700-01-01 00:00:00", tz = "UTC") + days(0:(600 * 365 + 145)) 
range(times)
```


```{r, warning=FALSE}
meeus_dn.df <- day_night(times, tz = "UTC", unit.out = "datetime", twilight = "sunlight")
meeus_dn.df
```

Now we use the 'suncalc' package to compute the same values.

```{r}
suncalc_dn.df <- getSunlightTimes(date = as.Date(times, tz = "UTC"), lat = 51.5, lon = 0)
head(suncalc_dn.df, n = 3L)
```

We compute the differences in sun elevations, and their range. Over this
extended period, the maximum and minimum differences are substantial. We plot a
smooth spline on the absolute value of the differences, as from the range we
already know that they maxima and minimum differences are of opposite signs. The
sun elevation values from the two functions are closest during the first half of
the 21st century. On the other hand we see that the fitted spline does not reach
the extreme values by a large margin. What is going on?

```{r, message=FALSE}
noon_differences <- suncalc_dn.df$solarNoon - meeus_dn.df$noon
range(differences)

ggplot(data.frame(times, noon_differences), aes(times, noon_differences)) +
  stat_hdr() +
  stat_panel_counts(label.x = "centre") +
  labs(x = "Time", y = "Difference in noon time (seconds)")
```

```{r, message=FALSE}
sunrise_differences <- suncalc_dn.df$sunrise - meeus_dn.df$sunrise
range(sunrise_differences)

ggplot(data.frame(times, sunrise_differences), aes(times, sunrise_differences)) +
  stat_hdr() +
  stat_panel_counts(label.x = "centre") +
  labs(x = "Time", y = "Difference in sunrise time (seconds)")
```

So do these shifts affect the day length estimates?

```{r, message=FALSE}

suncalc_dn.df$daylength <- suncalc_dn.df$sunset - suncalc_dn.df$sunrise

daylength_differences <- suncalc_dn.df$daylength - meeus_dn.df$daylength
range(daylength_differences)

ggplot(data.frame(times, daylength_differences), aes(times, daylength_differences)) +
  stat_hdr() +
  stat_panel_counts(label.x = "centre") +
  labs(x = "Time", y = "Difference in daylength (h)")
```

The differences seem to be consistently increasing as we move away from the 
first third of the 21st century, 2016 or 2000 seem to be the least different
depending on the value estimated.

## Conclusion

This tests confirm that the computation intensive algorithm from Meeus as
implemented in package 'SunCalcMeeus' provides much better estimates of sunset
and sunrise times, and of the position of the sun than package 'suncalc' at the
expense of slower performance that is in most cases tolerable. The improvement
in accuracy of using 'SunCalcMeeus' vs. 'suncalc' increases as time are farther
away from the present. Even if computations take nearly 3 times as long, 
they are still fast.

Tests for dates further into the past would be needed. Obtaining reference data
for a comparison requires more effort.

[On-line documentation](https://docs.r4photobiology.info/SunCalcMeeus/).
[NEWS file](https://docs.r4photobiology.info/SunCalcMeeus/news/).

